---
title: "Data Viz"
author: "Jake Riley"
date: "June 8, 2019"
output: ioslides_presentation
---

## TODO, eval = FALSE
 - export to powerpoint
 - explain pipe highlight that you pipe into ggplot
 - exlain that I'm not always going to use the argument names
 - preface that audience is for folks familiar with ggplot and tidyverse
 - put answers at the end
 - swap the first two exercises
 - show plot first in facet
 - show facet_wrap with scales = free
 - move golden ratio to best practices
 - remove coord_flip from boxplot - move to Appendix
 - retitle: Take care when cropping data
 - title for "Group A in the lead" shows wrong line

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  out.height = "300px"
)

library(tidyverse)
```

## today's talk
* https://bit.ly/2Wt7nkQ
* Intro
* What is ggplot
* Tips & Tricks
* Best Practices
* Try it out

## an intro

* Jake Riley
* Clinical Data Analyst at Children's Hospital of Philadelphia
* Avid `ggplot2` answerer on stackoverflow
* dogdad

@yake_84

https://bit.ly/2Wt7nkQ

## what is ggplot?
 * grammar of graphics
 * just like every sentence has a **subject, verb, and noun**, every chart has a **coordinate system, geom, and aesthetics**
 * the hope is that we will invent new types of charts
 
```{r}
p <- 
  ggplot(mpg) + 
  geom_count(
    aes(cty, hwy, color = class), 
    alpha = 0.8
  ) +
  theme_bw()

```
 
 
## an example
```{r grammar, out.width = c("50%", "50%"), fig.show = "hold"}
p

p + coord_polar()
```

## demystifying aes()

* `aes()` = aesthetics
* dynamic, data driven **variables** go inside the `aes()`
* constant, static **values** go outside

## note the difference
* `geom_point(aes(color = class, size = n), ...)`
* `geom_point(aes(...), color = "blue", size = 2)`

```{r out.width = c("50%", "50%"), fig.show = "hold", echo = FALSE}
df <-
  mpg %>% 
  count(cyl = factor(cyl), cty, hwy)

ggplot(df, aes(x = cty, y = hwy)) +
  geom_point(aes(color = cyl, size = n))

ggplot(df, aes(x = cty, y = hwy)) +
  geom_point(color = "blue", size = 2)

```


# Tips & Tricks

## descending bar charts

The number one things I get asked is how to make a barchart go in descending order. 

```{r}
ggplot(mpg, aes(class)) + 
  geom_bar() + 
  coord_flip()
```

## arrange by volume: `fct_infreq()`
```{r}
mpg %>%
  mutate(class = fct_infreq(class)) %>% 
  ggplot(aes(class)) + 
  geom_bar() + 
  coord_flip()
```

## arrange in descending order: `fct_rev()`
```{r, }
mpg %>% 
  mutate(class = fct_infreq(class)) %>% 
  ggplot(aes(fct_rev(class))) + 
  geom_bar() + 
  coord_flip()
```

## aggregated data: `fct_reorder()`
```{r}
mpg %>% 
  count(class) %>% 
  mutate(class = fct_reorder(class, n, sum))
```

## aggregated data: `fct_reorder()`
```{r}
mpg %>% 
  count(class) %>% 
  mutate(class = fct_reorder(class, n, sum)) %>% 
  ggplot(aes(class, n)) +
  geom_col() + 
  coord_flip()
```

## can you put this in descending order?
*(bonus) Can you do it without using a method I already mentioned?*
```{r}
mpg %>% 
  count(class) %>% 
  mutate(class = fct_reorder(class, n, sum)) %>% 
  ggplot(aes(class, n)) +
  geom_col() + 
  coord_flip()
```

## too many bars
```{r}
ggplot(mpg, aes(model)) +
  geom_bar() +
  coord_flip()
```

## too many bars: `fct_lump()`
```{r}
mpg %>%
  mutate(
    manufacturer = fct_lump(manufacturer, 5),
    manufacturer = fct_infreq(manufacturer),
    manufacturer = fct_rev(manufacturer)
  ) %>% 
  ggplot(aes(x = manufacturer)) +
  geom_bar() +
  coord_flip()
```

## order of fill
```{r}
ggplot(mpg, aes(manufacturer, fill = (cyl == 8))) +
  geom_bar() +
  coord_flip()
```

## order of fill
```{r}
mpg %>% 
  mutate(
    cyl_8 = (cyl == 8),
    manufacturer = fct_reorder(manufacturer, cyl_8, sum)
  ) %>% 
  ggplot(aes(manufacturer, fill = cyl_8)) +
  geom_bar() +
  coord_flip()
```

# can you show the top 3 classes ordered by cyl != 8?

# can you swap the `TRUE` & `FALSE` colors of the manufacturer plot?

## `facet_grid()` vs `facet_wrap()`
```{r}
p <-
  mpg %>% 
  filter(class != "2seater", cyl != 5) %>% 
  ggplot(aes(cty, hwy, color = class)) + 
  geom_count(alpha = 0.5) +
  lims(x = c(0, NA), y = c(0, NA)) + # can also use xlim() or scale_x_continuous
  guides(color = FALSE) +
  theme(aspect.ratio = 1)
```

## facet_grid: new syntax

```{r}
# this is the new syntax, replaces `facet_grid(~class)`
p + facet_grid(cols = vars(class))
```

## facets: scales

```{r}
# this is the old syntax, scales allows the x & y to vary
# also "free_x", "free_y"
p + facet_grid(cols = vars(class), scales = "free") 
```

## facets: scales & space

```{r}
p + facet_grid(cols = vars(class), scales = "free", space = "free")
```

## facets: margins

```{r}
p + 
  facet_grid(
    rows = vars(year),
    cols = vars(class),
    margins = TRUE
  )
```

## facet_wrap: # of columns/rows

```{r}
# also nrow
p + facet_wrap(~class, ncol = 3)
```

## facets: (a + b)

```{r}
# also works with facet_grid
p + facet_wrap(~class + year, nrow = 2)
```

## `scale_*_identity()`

Sometimes I want to have better control over colors & sizes. 

Here, I am hard coding the colors

```{r}
df <-
  mpg %>%
  mutate(category = 
    case_when(
      cty < 14 ~ "coral",
      cty > 19 ~ "turquoise",
      TRUE ~ "grey40"
    )
  )
```

## `scale_color_identity()`

```{r}
ggplot(df, aes(cty, hwy, color = category)) +
  geom_count() +
  scale_color_identity()
```

## `scale_fill_identity()`

```{r}

ggplot(df, aes(cty, fill = category)) +
  geom_dotplot() +
  scale_fill_identity()
```


## The golden ratio 1:1.6

* Try to give your charts the proportion of a credit card
* Also look this up

```{r, out.width = c("60%", "40%"), fig.show = "hold"}
p <- ggplot(mpg, aes(cty)) + geom_bar()

p + coord_fixed(1/10)
p + theme(aspect.ratio = 1/1.6) # ratio depends on the units
```


## Trouble with outliers
```{r, echo = FALSE}
p <- ggplot(diamonds, aes("", price)) + geom_boxplot()

find_limits <- function(x) {
  floor(ggplot_build(x)$data[[1]][2:4])
}

p + coord_flip()
```

## Use `coord_cartesian()` to zoom in

```{r}
# find_limits() is a custom function
bind_rows(
  find_limits(p),
  find_limits(p + ylim(0, 12000)),
  find_limits(p + scale_y_continuous(limits = c(0, 12000))),
  find_limits(p + coord_cartesian(ylim = c(0, 12000)))
)

```

# Best practices

## Dealing with spaghetti charts
```{r echo = FALSE}
set.seed(1234)
df <-
  expand.grid(
    group = LETTERS[1:5],
    x = as.Date(paste0("2019-01-", 1:30)),
    KEEP.OUT.ATTRS = FALSE
  ) %>% 
  mutate(
    rand = rnorm(-74:75, mean = 0, sd = 4),
    y = ((row_number()/2) %/% 3) + rand,
    category = factor(ntile(y, 4))
  )
  #mutate(y = row_number() %% 3 + row_number() %% 4 + row_number()/150)

my_theme <-
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    aspect.ratio = 5/30,
    legend.position = "bottom"
  )
```
This is one of the most common questions to answer: **change for multiple categories over time**. This often results in a chart like the one below. It is hard to read but there are some ways you can help your audience
```{r}
ggplot(df, aes(x, y, color = group)) + 
  geom_line()
```

## Highlight the focus & use an informative title
```{r}
ggplot(df, aes(x, y, group = group)) + 
  geom_line(data = filter(df, group != "A"), color = "grey", size = 1) +
  geom_line(data = filter(df, group == "A"), color = "black", size = 2) +
  labs(title = "Group A is currently in the lead")
```

## Try a heatmap but beware
[colorblind palette viewer](https://davidmathlogic.com/colorblind/#%2334804B-%236FD46B-%23D5D6D6-%23E5E6E6-%23B53D13)
```{r}
ggplot(df, aes(x, fct_reorder(group, y, last), fill = y)) + 
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "red", mid = "yellow", high = "darkgreen", midpoint = 12
  ) +
  my_theme +
  labs(title = "An improvement, but not colorblind friendly")
```


## Not every point needs a color

```{r}
ggplot(df, aes(x, fct_reorder(group, y), fill = category)) + 
  geom_tile(color = "white", size = 0.1) +
  geom_vline(xintercept = as.Date("2019-01-29")+0.5, size = 1.5) +
  scale_fill_manual(
    values = c("grey90", "grey85", "palegreen3", "seagreen4"),
    labels = c("poor", "good", "very good", "excellent")
  ) +
  my_theme +
  labs(title = "All teams now performing their best")
```


## Stackoverflow

* try `datapasta` for a minimal reprex 
* include images
* incorporate `styler`


# Extensions & Addins

## Extensions & Addins
* my `simplecolors` package
* [ggplot addins](https://github.com/daattali/addinslist)
* [ggplot extensions](https://www.ggplot2-exts.org/ggiraph.html)
* https://rjake.shinyapps.io/data_explorer/
* [R4DS](https://r4ds.had.co.nz/)


## `simplecolors`

https://rjake.github.io/simplecolors/articles/intro.html

```{r}
#devtools::install_github("rjake/simplecolors")
library(simplecolors)
simplecolors::show_colors(labels = FALSE)
```

## `simplecolors`

Three main functions:

* `sc()`
* `sc_across()`
* `sc_*()`

## `addinslist`:

`addinslist::addinslistAddin()`

```{r eval = FALSE, echo = TRUE}
# install.packages("addinslist")
# install.packages("esquisse")
# install.packages("ggedit")
# install.packages("ggThemeAssist")
# install.packages("colourpicker")

data(iris)
data(mpg)

p <-
  ggplot(mpg, aes(cty, hwy)) + 
  geom_point()
```

## Addins in action
```{r eval = FALSE, echo = TRUE}
#esquisse
  esquisse:::esquisser()
  esquisse:::esquisser(mpg)
  
# ggThemeAssist::ggThemeAssistGadget(p)
# ggedit(p)
# colourpicker
```

## Extensions

* [ggradar - spider/radar plots](https://github.com/ricardo-bion/ggradar)
* [gganimate](https://gganimate.com/)
* [ggrepel](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)
* [ggforce](https://cran.rstudio.com/web/packages/ggforce/vignettes/Visual_Guide.html)
* [cowplot](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html)
* [more](https://www.ggplot2-exts.org/ggiraph.html)

## Go Explore

* find a function or feature that you think is interesting or useful
* place screenshots here https://bit.ly/2XiG5C7
* we'll share 

